# Introduction

## What this course isn't, and is ##

Is Not: 

- statistics (in the way you think)
- general introduction to software engineering

Is: 

- data cleaning, data management, and exploratory data analysis
- habituation to some good software engineering practices that are especially valuable for data science work

## Notes on some design decisions ##

**R, not Python**

- Python is a perfectly good language for general programming
    - And has some advantages over R in certain data science aspects, including working with strings and developing custom data structures

- [R is a better language for doing data science](https://blog.shotwell.ca/posts/why_i_use_r/)
    - Vectors, data frames, and common statistical models are installed and loaded by default
    - R is more [functional](https://towardsdatascience.com/python-vs-r-choosing-the-best-programming-languages-for-data-science-b1327f01f6bf)
- R is also better for beginners
    - CRAN
    - RStudio
    
**Scripts, not notebooks**

- Quite a few scientists like the way [notebooks mix together code, visualizations, tables, and text](https://www.nature.com/articles/d41586-018-07196-1)

- [Notebooks do not play nicely with version control](https://towardsdatascience.com/version-control-with-jupyter-notebooks-f096f4d7035a)
- [Notebooks encourage bad habits and discourage good ones](https://docs.google.com/presentation/d/1n2RlMdmv1p25Xy5thJUhkKGvjtV-dkAIsUXP-AL4ffI/edit?usp=sharing)


## Let's get started with RStudio ##

* Quick tour of RStudio's UI
* Let's plot a thing

```{r}
data(anscombe) ## <https://en.wikipedia.org/wiki/Anscombe%27s_quartet>

anscombe
str(anscombe)
summary(anscombe)

library(ggplot2)
ggplot(data = anscombe, aes(x = x1, y = y1)) +
    geom_point()
```

**Exercise:** plot the other three datasets

```{r}
anscombe$x1
anscombe[['x1']]

anscombe[,'x1']
anscombe[,1]

mean(anscombe$x1)
mean(anscombe$x2)

sd(anscombe$y1)
sd(anscombe$y2)

cor(anscombe$x1, anscombe$y1)
```

**Exercise**: Calculate correlations for all four x-y pairs

* Some idiosyncracies of my code
    - `'` rather than `"`
    - `##` for documentation comment
    - `=` not `<-` for assignment

* Essential Global Options
    - Never save .Rdata, load workspace
* Super useful custom keyboard commands: 
    - Tools $\to$ Modify Keyboard Shortcuts...
    - ⌘+1, ⌘+2, ⌘+3

## Base R data types ##

### Numerics
```{r}
foo = 2
bar = 3
foobar = foo + bar
foo / bar
try(foo / 0)
```

### Vectors
```{r}
foo = c(2, 3, 4)
bar = c(5, 6, 7)
foo[2]
foo[2:3]
try(foo[1, 3])
foo[c(1, 3)]

foo + bar
foo * bar
## NB difference between vector and length-1 comparisons
bar == foo + 3
identical(bar, foo + 3)

c(foo, bar)
```

### Strings
```{r}
foo = c('monkey', 'banana')
bar = c('house')
## How to put these together? 
try(foo + bar)
c(foo, bar)
paste(foo, bar)
paste0(foo, bar)
```

### Factors
```{r}
foo_str = c('foo', 'foo', 'bar', 'zoo')
as.integer(foo_str)
foo_fct = as.factor(foo_str)
as.integer(foo)
levels(foo_fct)
```

### Lists 
*[todo]*

### Data frames
*[todo]*


### *[examining variables]*
- length
- str
- class, typeof
- [challenge](https://gge-ucd.github.io/R-DAVIS/lesson_how_r_thinks_about_data.html#challenge)

### *[Missing data]*

<https://gge-ucd.github.io/R-DAVIS/lesson_how_r_thinks_about_data.html#missing_data>