# (PART) Some fundamental tools {-}

# R Basics #

## Let's get started with RStudio ##

* Quick tour of RStudio's UI
* Let's plot a thing

```{r}
data(anscombe) ## <https://en.wikipedia.org/wiki/Anscombe%27s_quartet>

anscombe
str(anscombe)
summary(anscombe)

library(ggplot2)
ggplot(data = anscombe, aes(x = x1, y = y1)) +
    geom_point()
```

**Exercise:** plot the other three datasets

```{r}
anscombe$x1
anscombe[['x1']]

anscombe[,'x1']
anscombe[,1]

mean(anscombe$x1)
mean(anscombe$x2)

sd(anscombe$y1)
sd(anscombe$y2)

cor(anscombe$x1, anscombe$y1)
```

**Exercise**: Calculate correlations for all four x-y pairs

* Some idiosyncracies of my code
    - `'` rather than `"`
    - `##` for documentation comment
    - `=` not `<-` for assignment

* Essential Global Options
    - Never save .Rdata, load workspace
* Super useful custom keyboard commands: 
    - Tools $\to$ Modify Keyboard Shortcuts...
    - ⌘+1, ⌘+2, ⌘+3

## Base R data types ##

### Numerics
```{r}
foo = 2
bar = 3
foobar = foo + bar
foo / bar
try(foo / 0)
```

### Vectors
```{r}
foo = c(2, 3, 4)
bar = c(5, 6, 7)
foo[2]
foo[2:3]
try(foo[1, 3])
foo[c(1, 3)]

foo + bar
foo * bar
## NB difference between vector and length-1 comparisons
bar == foo + 3
identical(bar, foo + 3)

c(foo, bar)
```

### Strings
```{r}
foo = c('monkey', 'banana')
bar = c('house')
## How to put these together? 
try(foo + bar)
c(foo, bar)
paste(foo, bar)
paste0(foo, bar)
```

### Factors
```{r}
foo_str = c('foo', 'foo', 'bar', 'zoo')
as.integer(foo_str)
foo_fct = as.factor(foo_str)
as.integer(foo)
levels(foo_fct)
```

### Lists 
*[todo]*

### Data frames
*[todo]*


### *[examining variables]*
- length
- str
- class, typeof
- [challenge](https://gge-ucd.github.io/R-DAVIS/lesson_how_r_thinks_about_data.html#challenge)

### *[Missing data]*

<https://gge-ucd.github.io/R-DAVIS/lesson_how_r_thinks_about_data.html#missing_data>

# Git and Data Management #

# Functional Programming and Data Science #

# Warnings, Errors, and Getting Help #

