# Covid EDA 1 #

## </br>

```{r child = 'knitr_config.Rmd'}
```

- This EDA will start in class and continue on to the lab
- We're interested in *the role of social distancing in the July 2020 Covid wave in California*

## The standard narrative ##

The standard narrative about the July 2020 wave goes like this: 

- California had some of the first confirmed cases of Covid-19 in the US
- California was also the first state to institute a stay-at-home order and encourage social distancing
- These social distancing policies are why California did not experience the large first wave in March 2020 (contrast NYC)
- Starting in May 2020, these policies were relaxed and "lockdown fatigue" meant that people were ignoring them anyways
- This is why California did experience a more significant wave in July 2020

--- 

Our research question:  Is reduced social distancing (measured using cellphone tracking data) correlated with increased Covid-19 case counts 4 weeks later? 

</br>

In class, we'll just be looking at the case counts. 

## Reflexivity

(For time, you'll do these on your own as part of the lab.)

## Setup 

```{r, dependencies, message = FALSE, warning = FALSE}
library(tidyverse)
library(tidylog)
library(skimr)
library(visdat)

library(assertthat)

theme_set(theme_bw())
```

```{r, daily-diffs, echo = FALSE}
daily_new = function(x, order_var) {
    diff = x - dplyr::lag(x, order_by = order_var)
    return(diff)
}
```


## Data ##

To save some time in class, I've prepared a CSV that combines three datasets: 

- [Covid-19 case and death counts, at the county level, collected by the *New York Times*](https://github.com/nytimes/covid-19-data/blob/master/README.md)
- Estimated county populations in 2018, from the US Census' American Community Survey
- ["Mobility data," based on cellphone tracking, from Google](https://kjhealy.github.io/covmobility/reference/google_mobility.html)

The dataset has also been filtered down to California.  

## Get the data ##

- This week's lab on GitHub
    - <https://github.com/data-science-methods/lab_w06_covid>
- `data` -> `covid.csv` -> "Download" (right-click and copy)

```{r, covid-data}
covid_file = file.path('data', 'covid.csv')
covid_url = 'https://github.com/data-science-methods/lab_w06_covid/raw/main/data/covid.csv'
if (!file.exists(covid_file)) {
    download.file(covid_url, covid_file)
}
## Original version
# covid_df = read_csv(covid_file, show_col_types = FALSE)
## With daily change
covid_df = read_csv(covid_file, show_col_types = FALSE) |> 
    group_by(county) |>
    mutate(across(.cols = c(cases, deaths),
                  .fns = list(new = daily_new), date)) |> 
    ungroup()
```

## Some quick data checking ##

```{r}
skim(covid_df)
```

## Variables ##

- `county`, `state`, and `fips` are all geographic identifiers, and complete
    - 58 unique values for both `county` and `fips`
    - How many counties does California have? 
- `date` gives the date, is also complete
- `cases` and `deaths` come from the NYT covid data
- `population` comes from the Census
- `retail`, `grocery`, `parks`, `transit`, `workplaces`, and `residential` come from the Google Mobility data
    - Much lower completion rate

## Number of observations per county ##

```{r}
ggplot(covid_df, aes(fct_rev(fct_infreq(county)))) +
    # geom_bar() +
    geom_point(stat = 'count') +
    geom_segment(stat = 'count', 
                 aes(xend = county),
                 yend = 0) +
    coord_flip() +
    labs(x = 'county')
```

*Why do different counties have different numbers of observations?*


## No more than 1 observation per county per day ##

```{r}
covid_df |> 
    count(county, date) |> 
    filter(n > 1) |> 
    nrow() |> 
    identical(0L) |> 
    assert_that(msg = 'More than 1 observation per county per day')
```


## Cumulative vs. daily cases ##

Both `cases` and `deaths` are cumulative, not the daily new value

```{r}
covid_df |> 
    filter(county == 'San Francisco') |> 
    ggplot(aes(date, cases)) +
    geom_line()
```

---

- We'll write a little function to calculate the differences
- Incorporate it into the pipe that loads the data
    - Then restart R and rerun

```{r, daily-diffs, echo = TRUE, eval = FALSE}
```

---

- Test it on Orange and LA Counties

```{r, eval = FALSE}
covid_df |>
    filter(county %in% c('Orange', 'Los Angeles')) |>
    group_by(county) |>
    mutate(across(.cols = c(cases, deaths),
                  .fns = list(new = daily_new), date)) |>
    select(date, county, matches('cases'), matches('deaths')) |> 
    View()
```

---

Now we have daily values

```{r}
covid_df |> 
    filter(county == 'San Francisco') |> 
    ggplot(aes(date, cases_new)) +
    geom_line()
```

## Add an assertion ##

- For each county, the first daily diff should be `NA`
- But none of the others

```{r}
covid_df |> 
    group_by(county) |> 
    slice(-1) |> 
    pull(cases_new) |> 
    is.na() |> 
    any() |> 
    magrittr::not() |> 
    assert_that(msg = 'missing values in cases_new')
```

## Distribution of cases by county ##

First pass is hard to read

```{r}
ggplot(covid_df, aes(county, cases_new)) +
    geom_boxplot()
```

## Revisions

- `coef = 1000`
- y-axis on a log scale
- flip the coordinates
- reorder the counties by median number of cases
- filter out `cases_new == 0`
- meaningful axis labels

```{r}
covid_df |> 
    filter(cases_new > 0) |>
    ggplot(aes(fct_reorder(county, cases_new, 
                           .fun = median, na.rm = TRUE),
                     cases_new)) +
    geom_boxplot(coef = 1000) +
    scale_y_log10() +
    coord_flip() +
    labs(x = 'county', y = 'daily new cases')
```

## 9 plots of the same plot ##

```{r}
ggplot(covid_df, aes(x = cases_new, y = deaths_new)) +
    # geom_point() +
    # geom_point(alpha = .2) +
    # geom_count(alpha = .5) +
    # geom_bin2d() +
    # geom_hex() +
    # geom_hex(aes(color = after_stat(count))) +
    # geom_density2d(size = 1) +
    # stat_density2d(contour = FALSE, geom = 'raster',
    #                aes(fill = after_stat(density)),
    #                show.legend = FALSE) +
    stat_density2d(contour = TRUE, geom = 'polygon',
                   aes(fill = after_stat(level)),
                   show.legend = FALSE) +
    scale_x_log10() +
    scale_y_log10() +
    scale_color_viridis_c(aesthetics = c('color', 'fill'))
```


