# Readable code is reliable code #

```{r child = 'knitr_config.Rmd'}
```

## Reading 

- @BryanCodeSmellsFeels2018
- @PostolovskiYourCodeReview2020
- @BuckensSelfdocumentingMythHow2019


## Writing code is writing

- Multiple target audiences
    - Collaborators
    - (Some) reviewers and readers of the paper
    - Peers who want to analyze and extend your methods
    - *Yourself in two months/one year/five years*

- Your code is *readable* to the extent that people can use it to easily and reliably 
    - predict, 
    - diagnose, and
    - extend your code


## Code style ##

- iNéwritteNélanguagEáconventionSéfoRîpunctuatioNøcapitalizatioNîaiDécomprehensioNébYéindicatinGéstructurE
    - this is what it's like to read poorly-styled code
    - conventions only work if they're *shared* conventions

- Style guides provide shared conventions for readable code
    - In-line spacing makes it easier to pick out distinguish functions, operators, and variables in a line
    - Returns distinguish arguments in a function call
    - Indentation corresponds to structure of complex expressions
    - Common conventions for naming, assignment reduce cognitive load

- Tidyverse style guide:  <https://style.tidyverse.org/>


## Highlights from the Tidyverse style guide ##

- Place all `package()` calls at the top of the script
- Limit your code to 80 characters per line
- Use at least 4 spaces for indenting multiline expression
    - Control-I in RStudio will do automagic indenting
- In multiline function calls, 1 argument = 1 line
- Use comments to explain "why" rather than "what" or "how"
    - DH's comment convention: 
    - Single `#` is for code that isn't currently used but might need to be brought back (eg, for debugging)
    - `##` is for substantial comments


## Spaces
- Always put spaces after commas, and never before (like English)
- But not between a function name and the parentheses (like math)
- And surrounding infix operators (`==`, `+`, `-`, `<-`, `=`)
- Pipes `%>%` `|>` should have a space before and be at the end of the line


## Code blocks
When you put a block of code in curly braces `{}`: 

- `{` should be the last character on a line
- `}` should be the first character on the line

```{r, eval = FALSE}
if (y == 0) {
  if (x > 0) {
    log(x)
  } else {
    message("x is negative or zero")
  }
} else {
  y^x
}
```


## Boolean variables vs. control flow

- Functions that return vectors: 
    - `&`, `|`, `==`, `ifelse()`, `dplyr::if_else()`
- Functions that return a single value: 
    - `&&`, `||`, `identical`
- `if (x) a else b` only looks at the first (hopefully single) value of `x`
