--- 
title: "Methods of Data Science I"
author: "Dan Hicks, <hicks.daniel.j@gmail.com>"
github-repo: data-science-methods/book
date: "`r date()`"
site: bookdown::bookdown_site
# documentclass: book
bibliography: [book.bib, packages.bib]
link-citations: yes
description: "This document is a set of notes (and perhaps later a textbook) for a course on data science methods for graduate students in social and behavioral sciences."
---

```{r child = 'knitr_config.Rmd'}
```


# Preface # {-} 

This document is a set of notes (and perhaps later a textbook) for a course on data science methods for graduate students in social and behavioral sciences, taught by Professor Dan Hicks (they/them), UC Merced. 

The course site is available [here](..). 

<!-- PDF versions of each chapter can be found here:  [Handouts](handouts/index.html) -->

<!-- (The URLs to the handouts are buggy.  To get to them, look for `handout/handout/` in the URL; delete the second `handout/`.) -->

-----

These notes were written in [RStudio](http://www.rstudio.com/ide/) using [bookdown](http://bookdown.org/). The complete source is available on [GitHub](https://github.com/data-science-methods/book/) and is automatically rebuilt using GitHub actions. 

This version of the book was built with `r R.version.string`, [pandoc](https://pandoc.org/) version `r rmarkdown::pandoc_version()`, and the following packages:


```{r pkg-list, echo = FALSE, results="asis"}
## From <https://github.com/tidymodels/TMwR/blob/master/index.Rmd>
deps <- desc::desc_get_deps()
pkgs <- sort(deps$package[deps$type == "Imports"])
pkgs <- sessioninfo::package_info(pkgs, dependencies = FALSE)
df <- tibble::tibble(
  package = pkgs$package,
  version = pkgs$ondiskversion,
  source = gsub("@", "\\\\@", pkgs$source)
)
knitr::kable(df, format = "markdown")
```
