# Visual EDA #

- Recommended reading: @WeissgerberBarLineGraphs2015; @WickhamLayeredGrammarGraphics2010
- Further reading: @HealyDataVisualizationPractical2018, ch. 1; @MatejkaSameStatsDifferent

```{r child = 'knitr_config.Rmd'}
```

## The Role of Visuals in EDA ##

- @TukeyExploratoryDataAnalysis1977 emphasized visual methods for EDA
- Including not just graphs but also structured tables, such as [stem-and-leaf displays](https://en.wikipedia.org/wiki/Stem-and-leaf_display)

- In terms of our 3 models of EDA (ch. \@ref(eda-models))
    - Risk management:  With the right plot, our visual system can very efficiently detect potential problems in the data that don't show up easily with summary statistics
        - Outliers 
        - Degeneracies (eg, perfect correlation)
        - Skew or bimodal distributions
        - Non-linear relationships

    - Epicycle of analysis:  
        - Check expectations about the distribution of variables
        - Develop expectations about relationships between variables

    - Phenomena development: 
        - Quickly identify potential patterns
        - Contrast potential patterns with noise/uncertainty/imprecision


## Plot Your Data ##

- Summary statistics almost always focus only on central tendency (mean, median) and dispersion (standard deviation, IQR)
    - This is all you would need if the world were made of normal distributions
    - The world is not made of normal distributions [@LyonWhyAreNormal2014]

- We'll illustrate this using The Datasaurus Dozen, a collection of 13 datasets [@MatejkaSameStatsDifferent]
    - Each dataset has two variables, `x` and `y`

```{r}
library(tidyverse)
library(datasauRus)
dataf = datasaurus_dozen

dataf
count(dataf, dataset)
```

- The datasets have the same means, standard deviations, and (Pearson) correlation coefficient
    - p-value of the correlation coefficient is not statistically significant

```{r}
## This is a more complex summarize() call than we've seen before
## 1. Number of rows
## 2. "Summarize across the columns x and y, using the functions mean and sd"; automatically generates names
## 3. Correlation between x and y
## 4. p-value from a t-test of the null that the correlation = 0
dataf %>% 
    group_by(dataset) %>% 
    summarize(n = n(), 
              across(.cols = c(x, y), 
                     .fns = lst(mean, sd)), 
              cor_xy = cor(x, y), 
              p_value = cor.test(x, y)$p.value) %>% 
    ungroup()
```

```{r}
ggplot(dataf, aes(x, y)) +
    geom_point() +
    facet_wrap(vars(dataset))
```

